The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-30 paper/paper-jsdse/paper-jsdse.tex
navigation step changed to figure

\documentclass[12pt]{article}
\usepackage{amsmath}
%\usepackage{longtable}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{booktabs}
\usepackage{url} % not crucial - just used below for the URL
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

The NLSY79 data used in Singer and Willett (2003). contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994. In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years. Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.

@@

M. van der Loo and de Jonge (2018) describe the notion of a ``statistical value chain'' where the production stages of data cleaning process are earmarked as raw data (data from the data source), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables). In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidying the raw data into input data, specifically for wages data (Section \ref{tidywages}), demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida}.

@@

The NLSY79 data contains a large scope of variables but for our aim, the scope required is limited to demographic profiles and wages data. More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described below.

\begin{tcolorbox}[label="box:data", title=Navigating the data source to download the raw data]

\faDatabase~NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck~The CASEID will be always be selected.\\
\vspace{1mm}
\faCheck~The 3 recommended demographic variable (sample ID, race and sex) were selected.\\
\vspace{1mm}
For the remaining variables, we went to the ``Variable Search'' tab and select variables as follows

\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 442 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 686 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}

\end{tcolorbox}

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
%\usepackage{longtable}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{booktabs}
\usepackage{float}
\usepackage{url} % not crucial - just used below for the URL
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

The NLSY79 data used in Singer and Willett (2003) contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994. In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years. Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.

@@

M. van der Loo and de Jonge (2018) describe the notion of a ``statistical value chain'' where the production stages of data cleaning process are earmarked as raw data (data as arrived to the desk of an analyst), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables). In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidying the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida}.

@@

The NLSY79 data contains a large scope of variables but for our aim, the scope required is limited to demographic profiles and wages data. More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described in Figure \ref{fig:source-nav}.

\begin{figure}[Ht]

\begin{tcolorbox}
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck\ The CASEID will be always be selected.  \\
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
\vspace{1mm}
For the remaining variables, we went to the "Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 442 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 686 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}
\end{tcolorbox}
\caption{Above shows the steps taken to navigate the data source to download the raw data.\label{fig:source-nav}}
\end{figure}

### 2021-09-30 paper/paper.Rmd
navigation step changed to figure

The NLSY79 data used in @SingerJudithD2003Alda. contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994. In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years. Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.

@@

@LooMarkvander2018Sdcw describe the notion of a "statistical value chain" where the production stages of data cleaning process are earmarked as raw data (data from the data source), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables).  In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidying the raw data into input data, specifically for wages data (Section \@ref(tidywages)), demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida).

@@

The NLSY79 data contains a large scope of variables but for our aim, the scope required is limited to demographic profiles and wages data. More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described below.

::: {.tcolorbox data-latex="[label=\"box:data\", title=Navigating the data source to download the raw data]" }
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})  
\vspace{1mm}
\faCheck\ The CASEID will be always be selected.  
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  
\vspace{1mm}
For the remaining variables, we went to the "Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 442 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 686 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}
:::

>>>>

The NLSY79 data used in @SingerJudithD2003Alda contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994. In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years. Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.

@@

@LooMarkvander2018Sdcw describe the notion of a "statistical value chain" where the production stages of data cleaning process are earmarked as raw data (data as arrived to the desk of an analyst), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables).  In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidying the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida).

@@

The NLSY79 data contains a large scope of variables but for our aim, the scope required is limited to demographic profiles and wages data. More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described in Figure \ref{fig:source-nav}.

\begin{figure}[t]

\begin{tcolorbox}[title = Navigating the data source]
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck\ The CASEID will be always be selected.  \\
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
\vspace{1mm}
For the remaining variables, we went to the "Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 442 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 686 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}
\end{tcolorbox}
\caption{The above documents the steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
\end{figure}

### 2021-09-30 paper/paper.tex
navigation step changed to figure

\begin{abstract}
The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data is never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets, and the code to produce them are provided in an accompanying open source R package, called \texttt{[CENSORED]}. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks, and providing more convenient samples of the data to potential users.
\end{abstract}

@@

The NLSY79 data contains a large scope of variables but for our aim, the scope required is limited to demographic profiles and wages data. More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described below.

\begin{tcolorbox}[label="box:data", title=Navigating the data source to download the raw data]

\faDatabase~NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck~The CASEID will be always be selected.\\
\vspace{1mm}
\faCheck~The 3 recommended demographic variable (sample ID, race and sex) were selected.\\
\vspace{1mm}
For the remaining variables, we went to the ``Variable Search'' tab and select variables as follows

\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 442 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 686 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}

\end{tcolorbox}

>>>>

\begin{abstract}
The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data is never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets, and the code to produce them are provided in an accompanying open source R package, called {[}CENSORED{]}. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks, and providing more convenient samples of the data to potential users.
\end{abstract}

@@

The NLSY79 data contains a large scope of variables but for our aim, the scope required is limited to demographic profiles and wages data. More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described in Figure \ref{fig:source-nav}.

\begin{figure}[t]

\begin{tcolorbox}[title = Navigating the data source]
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck\ The CASEID will be always be selected.  \\
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
\vspace{1mm}
For the remaining variables, we went to the "Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 442 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 686 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}
\end{tcolorbox}
\caption{The above documents the steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
\end{figure}
